---
title: "High-frequency seasonal adjustment with rjd3highfreq"
author: "Giancarlo Lutero (ISTAT)"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{High-frequency seasonal adjustment with rjd3highfreq}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4
)
```

```{r libraries}
options(java.parameters = "-Xmx8g") # High RAM usage expected
library(rjd3toolkit)
library(rjd3highfreq)
```

## Introduction

The `rjd3highfreq` package provides R wrappers to the JD+ Java 
implementation for seasonal adjustment and decomposition of
high-frequency time series such as daily data.

This vignette illustrates a complete workflow including fractional
airline estimation, linearization, single- and multi-period
decompositions, and visualization.

## Synthetic example

```{r synthetic-data}
set.seed(123)

data <- list(
  y = rnorm(730, mean = 100, sd = 5),
  date = seq.Date(
    from = as.Date("2022-01-01"),
    by = "day",
    length.out = 730
  )
)
```

```{r fractional-airline}
est <- fractionalAirlineEstimation(
  y = data$y,
  periods = c(7, 365.2425),
  log = FALSE,
  y_time = data$date
)

linearized_data <- list(
  y = est$model$linearized,
  date = data$date
)
```

```{r multi-period-decomposition}
jdec <- multiAirlineDecomposition_raw(
  y = linearized_data$y,
  periods = c(7, 365.2425),
  stde = TRUE
)

amb_r <- jd2r_multiAirlineDecomposition(
  jrslt = jdec,
  stde = TRUE,
  periods = c(7, 365.2425),
  y_time = linearized_data$date
)
```

```{r plots-synthetic}
plot(est)
plot(amb_r, type_chart = "y-sa-trend")
```

```{r plots-larger,fig.width=7.5,fig.height=4}
# plot(amb_r, type_chart = "cal-seas-irr")
```

## Daily births in France
```{r births-data}
df_daily <- rjd3toolkit::Births
df_daily$date <- as.Date(df_daily$date)
```

```{r calendar}
frenchCalendar <- national_calendar(days = list(
  fixed_day(7, 14),
  fixed_day(5, 8, validity = list(start = "1982-05-08")),
  special_day("NEWYEAR"),
  special_day("MAYDAY"),
  special_day("EASTERMONDAY"),
  special_day("ASCENSION"),
  special_day("WHITMONDAY"),
  special_day("ASSUMPTION"),
  special_day("ALLSAINTSDAY"),
  special_day("ARMISTICE"),
  special_day("CHRISTMAS")
))
```

```{r calendar-regressors}
cal_reg <- holidays(
  calendar = frenchCalendar,
  start = "1968-01-01",
  length = nrow(df_daily),
  type = "All",
  nonworking = 7L
)
```

```{r births-estimation}
pre_pro <- fractionalAirlineEstimation(
  y = df_daily$births,
  x = cal_reg,
  periods = 7,
  outliers = c("ao", "wo"),
  log = TRUE,
  y_time = df_daily$date
)
```

```{r single-period-decompositions}
amb.dow <- fractionalAirlineDecomposition(
  y = pre_pro$model$linearized,
  period = 7,
  log = TRUE,
  y_time = df_daily$date
)

amb.doy <- fractionalAirlineDecomposition(
  y = amb.dow$decomposition$sa,
  period = 365.2425,
  log = TRUE,
  y_time = df_daily$date
)
```

```{r multi-period-births}
amb.multi <- multiAirlineDecomposition(
  y = pre_pro$model$linearized,
  periods = c(7, 365.2425),
  log = TRUE,
  y_time = df_daily$date
)

# plot(amb.multi)
```

## Conclusions

The `rjd3highfreq` package enables a coherent treatment of
high-frequency time series within the JD+ ecosystem, combining calendar
effects, fractional airline models, and flexible multi-period seasonal
decompositions.
